Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-06-12 LICS_2025/e_graphs_with_bindings/sections/combinatorial_semantics.tex
notes for slotted e-graphs and arxiv script

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
$
n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
$,
computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.
To make such cospans complete with respect to representing morphism of $\catname{S}(\Sigma)$, we need to impose some restriction on their form.
\begin{definition}[Degree,  path,  acyclic] 
The \emph{in-degree} of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.  
Similarly, the out-degree of $v$ is the number of edges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
A \emph{path} from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
A hypergraph is \emph{directed acyclic} if it contains no paths of length $n > 0$ that start and end in the same edge.
\end{definition}

@@

Practical meaning of the above theorem is that two $\Sigma$-terms $f$ and $g$ are equal modulo SMC equations if and only if they are represented as the same hypergraph with interfaces.
The theorem defines a full and faithful functor that is induced by interpretation function of morphism terms that renders the latter as cospans of hypergraphs and is given in Appendix~\ref{sec:appendix:interpretation}.
Term rewriting and hence equality modulo $\mathcal{E}$ is then implemented as DPO rewrites over such hypergraphs with interfaces that we recall next.

@@

We follow an analogous approach by combining combinatorial representations of~\cite{fscd} and~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
We next extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} to accommodate new structure brought by closedness, and when evoking a term e-hypergraph modify the original definitions as follows.

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{e-color}{<}, \textcolor{closed-color}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$ is a set of hyperedges that that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{e-color}{<} \subseteq \textcolor{e-color}{E} \times V + E$ and $\textcolor{closed-color}{<} \subseteq \textcolor{closed-color}{E} \times V + E $ with $\textcolor{e-color}{E}, \textcolor{closed-color} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{e-color}{<^{\mu}}$ and $\textcolor{closed-color}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{e-color}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{closed-color}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{e-color}{<^{\mu}} x$ or $e \textcolor{e-color}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{e-color}{<^{\mu}} e$ iff $e' \textcolor{e-color}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{closed-color}{<}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{e-color}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{e-color}{<^{\mu}}(v)$ and $p \textcolor{e-color}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{e-color}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{e-color}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{e-color}{E}$ and $\textcolor{closed-color}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{e-color}{<}, \textcolor{closed-color}{<}$ and~$\consistency$.
\end{definition}

@@

\begin{definition}[Isomorphic cospans]
Consider a relation 
\begin{multline*}
R = \bigl\{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y)\\ \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{closed-color}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{closed-color}{<}^{\mu} f_{\text{int}}(y)\bigr\}
\end{multline*}
for $x, y \in n'$.
And let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}^{k}_{j=1}$.
We get an analogous partition for $m'$.

@@

    \item for all $v,w$ in $\mathcal{G}$ in the image of $i + j$,  $v$ and $w$ share the same sequence of predecessors, and either $v,w$ are top-level, or $v \consistency w$, or there exists $e \in \textcolor{closed-color}{E}$ s.t. $v,w \textcolor{closed-color}{<}^{\mu} e$;
    \item for all $v,w$ in $\mathcal{L}^{\bot}$ in the image of $i + j$,  $v$ and $w$ share the same sequence of predecessors, and either $v,w$ are top-level or or $v \consistency w$, or there exists $e \in \textcolor{closed-color}{E}$ s.t. $v,w \textcolor{closed-color}{<}^{\mu} e$;

@@

@@

>>>>

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
$
n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
$,
computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.
To make such cospans complete with respect to representing morphisms of $\catname{S}(\Sigma)$, we need to impose some restrictions on their form.
\begin{definition}[Degree,  path,  acyclic] 
The \emph{in-degree} of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.  
Similarly, the out-degree of $v$ is the number of edges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
A \emph{path} from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
A hypergraph is \emph{directed acyclic} if it contains no paths of length $n > 0$ that start and end in the same edge.
\end{definition}

@@

The practical meaning of the above theorem is that two $\Sigma$-terms $f$ and $g$ are equal modulo SMC equations if and only if they are represented as the same hypergraph with interfaces.
The theorem defines a full and faithful functor that is induced by the interpretation function of morphism terms that renders the latter as cospans of hypergraphs and is given in Appendix~\ref{sec:appendix:interpretation}.
Term rewriting and hence equality modulo $\mathcal{E}$ is then implemented as DPO rewrites over such hypergraphs with interfaces that we recall next.

@@

We follow an analogous approach by combining combinatorial representations of~\cite{fscd} and~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
We next extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} to accommodate new structure brought by closedness, and when evoking a term `e-hypergraph' we will mean the structure defined as follows.

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{e-color}{<}, \textcolor{closed-color}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$ is a set of hyperedges that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{e-color}{<} \subseteq \textcolor{e-color}{E} \times V + E$ and $\textcolor{closed-color}{<} \subseteq \textcolor{closed-color}{E} \times V + E $ with $\textcolor{e-color}{E}, \textcolor{closed-color} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{e-color}{<^{\mu}}$ and $\textcolor{closed-color}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{e-color}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{closed-color}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{e-color}{<^{\mu}} x$ or $e \textcolor{e-color}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{e-color}{<^{\mu}} e$ iff $e' \textcolor{e-color}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{closed-color}{<}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{e-color}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{e-color}{<^{\mu}}(v)$ and $p \textcolor{e-color}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{e-color}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{e-color}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{e-color}{E}$ and $\textcolor{closed-color}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{e-color}{<}, \textcolor{closed-color}{<}$ and~$\consistency$.
\end{definition}

@@

\begin{definition}[Isomorphic cospans]
Consider a relation
\ifdefined\ONECOLUMN
\[
    R = \bigl\{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y) \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{closed-color}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{closed-color}{<^{\mu}} f_{\text{int}}(y)\bigr\}
\]
\else 
\begin{multline*}
R = \bigl\{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y)\\ \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{closed-color}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{closed-color}{<^{\mu}} f_{\text{int}}(y)\bigr\}
\end{multline*}
\fi
for $x, y \in n'$.
And let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}^{k}_{j=1}$.
We get an analogous partition for $m'$.

@@

    \item for all $v,w$ in $\mathcal{G}$ in the image of $i + j$,  $v$ and $w$ share the same sequence of predecessors, and either $v,w$ are top-level, or $v \consistency w$, or there exists $e \in \textcolor{closed-color}{E}$ s.t. $v,w \textcolor{closed-color}{<^{\mu}} e$;
    \item for all $v,w$ in $\mathcal{L}^{\bot}$ in the image of $i + j$,  $v$ and $w$ share the same sequence of predecessors, and either $v,w$ are top-level or or $v \consistency w$, or there exists $e \in \textcolor{closed-color}{E}$ s.t. $v,w \textcolor{closed-color}{<^{\mu}} e$;

@@

    \ifdefined\ONECOLUMN
        n \xrightarrow{f_1} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2} k
    \else

@@

    \fi

### 2025-06-12 LICS_2025/monoidal_e_graphs/sections/appendix.tex
notes for slotted e-graphs and arxiv script

            \tikzcdset{row sep/mysizerow/.initial=0.1ex}

@@

                n_2 + m_1 &&&&&&&& m_2 + n_1\\

@@

            \scalebox{0.75}{\tikzfig{../figures/appendix/contexts_3}}    

@@

\begin{figure*}
    \begin{subfigure}{0.4\linewidth}
    \[
    \adjustbox{scale=0.7}{
    \tikzfig{../figures/appendix/cartesian_rewrite_source_graph}
    }
    \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.55\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/cartesian_rewrite_dpo_stuck}
        }
        \]
    \end{subfigure}
    \caption{DPO rewriting stuck}
    \label{fig:dpo-stuck}
\end{figure*}
Another problematic rewrite arises when we want to utilise the structural rule of the free enrichment
\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.6\linewidth}{
\tikzfig{../figures/appendix/structural_rewrite_problem}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
\tikzfig{../figures/appendix/structural_rewrite_problem}
}
\]
\fi
if we consider the following DPO square
\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck}
}
\]
\fi

@@

where the labels show how interface edges are mapped to the edges of the corresponding e-hypergraphs.
For example, the edge in the interface labelled $a_1$ is mapped to the outermost edge on the right.
We can now compute the pushout complement and complete the DPO square below

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
}
\]
\fi

Handling deletion or introduction rewrite rules will require the introduction of auxiliary rules.
In particular, two rules for introduction and elimination of these special edges

\ifdefined\ONECOLUMN
\begin{figure}[h!]
    \begin{subfigure}{0.45\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/idI_elimination_hypergraph}
        }
        \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/idI_introduction_hypergraph}
        }
        \]
    \end{subfigure}
    \end{figure}
\else
\begin{figure}[h!]
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_elimination_hypergraph}
    }
    \]
\end{subfigure}
\hfill
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_introduction_hypergraph}
    }
    \]
\end{subfigure}
\end{figure}
\fi
and similar rules to introduce and eliminate in nested contexts (Figure~\ref{fig:nested_elimination});

\begin{figure}[h!]
    \begin{subfigure}{\linewidth}
        \[
        \adjustbox{scale=0.55}{
        \tikzfig{../figures/appendix/idI_elimination_nested_hypergraph}
        }
        \]
    \end{subfigure}
    \vspace{1em}
    \begin{subfigure}{\linewidth}
        \[
        \adjustbox{scale=0.55}{
        \tikzfig{../figures/appendix/idI_introduction_nested_hypergraph}
        }
        \]
    \end{subfigure}
    \caption{Special edge nested elimination}
    \label{fig:nested_elimination}
\end{figure}

and, finally, the rules to introduce such edges as new consistent components (Figure~\ref{fig:consistent_introduction}).
Intuitively such edge plays the role of a tensor unit identity that can be inserted in any context.

\ifdefined\ONECOLUMN
\begin{figure}[h!]
    \begin{subfigure}{0.6\linewidth}
        \[
        \adjustbox{width=\linewidth}{
        \tikzfig{../figures/appendix/idI_introduction_new_component_hypergraph}
        }
        \]
        \subcaption{}
    \end{subfigure}
    \begin{subfigure}{0.6\linewidth}
        \[
        \adjustbox{width=\linewidth}{
        \tikzfig{../figures/appendix/idI_elimination_new_component_hypergraph}
        }
        \]
        \subcaption{}
    \end{subfigure}
    \caption{New consistent component introduction (a) and elimination (b) rule}
    \label{fig:consistent_introduction}
\end{figure}
\else
\begin{figure}[h!]
    \begin{subfigure}{\linewidth}
        \[
        \adjustbox{width=\linewidth}{
        \tikzfig{../figures/appendix/idI_introduction_new_component_hypergraph}
        }
        \]
    \end{subfigure}
    \vspace{1em}
        \begin{subfigure}{\linewidth}
            \[
            \adjustbox{width=\linewidth}{
            \tikzfig{../figures/appendix/idI_elimination_new_component_hypergraph}
            }
            \]
        \end{subfigure}
    \caption{New consistent component introduction (top) and elimination (bottom) rule}
    \label{fig:consistent_introduction}
\end{figure}
\fi

@@

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.7\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix}
}
\]
\fi
The target e-hypergraph is not exactly what we wanted, however, but we can use the auxiliary rule to remove this unlabelled edge.

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.7\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix_2}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix_2}
}
\]
\fi
% and then utilise the idempotence rule for the identity of the tensor unit

>>>>

            \tikzcdset{row sep/mysizerow/.initial=0.3ex}

@@

                n_2 + m_1\hspace{-2em} &&&&&&&& \hspace{-2em}m_2 + n_1\\

@@

            \hspace{-1em}\scalebox{0.75}{\tikzfig{../figures/appendix/contexts_3}}    

@@

\begin{figure*}[t!]
    \begin{subfigure}{0.4\linewidth}
    \[
    \adjustbox{scale=0.7}{
    \tikzfig{../figures/appendix/cartesian_rewrite_source_graph}
    }
    \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.55\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/cartesian_rewrite_dpo_stuck}
        }
        \]
    \end{subfigure}
    \caption{DPO rewriting stuck}
    \label{fig:dpo-stuck}
\end{figure*}

\begin{figure*}[!t]
    \centering
    \begin{minipage}[t]{0.48\linewidth}
        \centering
            \begin{subfigure}{\linewidth}
                \[
                \adjustbox{width=0.95\linewidth}{
                \tikzfig{../figures/appendix/idI_elimination_nested_hypergraph}
                }
                \]
            \end{subfigure}
            \vspace{1em}
            \begin{subfigure}{\linewidth}
                \[
                \adjustbox{width=0.95\linewidth}{
                \tikzfig{../figures/appendix/idI_introduction_nested_hypergraph}
                }
                \]
            \end{subfigure}
            \caption{Special edge nested introduction (top) and elimination (bottom) rules}
            \label{fig:nested_elimination}
    \end{minipage}%
    \hfill
    \begin{minipage}[t]{0.48\linewidth}
        \centering
            \begin{subfigure}{\linewidth}
                \[
                \adjustbox{width=0.95\linewidth}{
                \tikzfig{../figures/appendix/idI_introduction_new_component_hypergraph}
                }
                \]
            \end{subfigure}
            \vspace{1em}
                \begin{subfigure}{\linewidth}
                    \[
                    \adjustbox{width=0.95\linewidth}{
                    \tikzfig{../figures/appendix/idI_elimination_new_component_hypergraph}
                    }
                    \]
                \end{subfigure}
            \caption{New consistent component introduction (top) and elimination (bottom) rules}
            \label{fig:consistent_introduction}        
    \end{minipage}
\end{figure*}
\begin{figure}[t!]
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
}
\]
\caption{Complete DPO square for free enrichment structural rule}
\label{fig:completed_dpo}
\end{figure}

Another problematic rewrite arises when we want to utilise the structural rule of the free enrichment
\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.6\linewidth}{
\tikzfig{../figures/appendix/structural_rewrite_problem}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
\tikzfig{../figures/appendix/structural_rewrite_problem}
}
\]
\fi
if we consider the following DPO square
\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck}
}
\]
\fi

@@

where the labels show how interface edges are mapped to the edges of the corresponding e-hypergraphs.
For example, the edge in the interface labelled $a_1$ is mapped to the outermost edge on the right.
We can now compute the pushout complement and complete the DPO square \ifdefined\ONECOLUMN below \else in Figure~\ref{fig:completed_dpo}. \fi

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
}
\]
\else
% \[
% \adjustbox{width=\linewidth}{
%     \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
% }
% \]
\fi

Handling deletion or introduction rewrite rules will require the introduction of auxiliary rules.
In particular, two rules for introduction and elimination of these special edges (Figure~\ref{fig:idI_intro_elim}),
\ifdefined\ONECOLUMN
\begin{figure}[h!]
    \begin{subfigure}{0.45\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/idI_elimination_hypergraph}
        }
        \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/idI_introduction_hypergraph}
        }
        \]
    \end{subfigure}
    \end{figure}
\else
\begin{figure}[h!]
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_elimination_hypergraph}
    }
    \]
\end{subfigure}
\hfill
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_introduction_hypergraph}
    }
    \]
\end{subfigure}
\caption{Special edge flat introduction and elimination}
\label{fig:idI_intro_elim}
\end{figure}
\fi
similar rules to introduce and eliminate in nested contexts (Figure~\ref{fig:nested_elimination});
% \begin{figure}[t!]
%     \begin{subfigure}{\linewidth}
%         \[
%         \adjustbox{scale=0.55}{
%         \tikzfig{../figures/appendix/idI_elimination_nested_hypergraph}
%         }
%         \]
%     \end{subfigure}
%     \vspace{1em}
%     \begin{subfigure}{\linewidth}
%         \[
%         \adjustbox{scale=0.55}{
%         \tikzfig{../figures/appendix/idI_introduction_nested_hypergraph}
%         }
%         \]
%     \end{subfigure}
%     \caption{Special edge nested elimination}
%     \label{fig:nested_elimination}
% \end{figure}
and, finally, the rules to introduce such edges as new consistent components (Figure~\ref{fig:consistent_introduction}).
Intuitively such edge plays the role of a tensor unit identity that can be inserted in any context.

\ifdefined\ONECOLUMN
\begin{figure}[h!]
    \begin{subfigure}{0.6\linewidth}
        \[
        \adjustbox{width=\linewidth}{
        \tikzfig{../figures/appendix/idI_introduction_new_component_hypergraph}
        }
        \]
        \subcaption{}
    \end{subfigure}
    \begin{subfigure}{0.6\linewidth}
        \[
        \adjustbox{width=\linewidth}{
        \tikzfig{../figures/appendix/idI_elimination_new_component_hypergraph}
        }
        \]
        \subcaption{}
    \end{subfigure}
    \caption{New consistent component introduction (a) and elimination (b) rule}
    \label{fig:consistent_introduction}
\end{figure}
\else
% \begin{figure}[t!]
%     \begin{subfigure}{\linewidth}
%         \[
%         \adjustbox{width=\linewidth}{
%         \tikzfig{../figures/appendix/idI_introduction_new_component_hypergraph}
%         }
%         \]
%     \end{subfigure}
%     \vspace{1em}
%         \begin{subfigure}{\linewidth}
%             \[
%             \adjustbox{width=\linewidth}{
%             \tikzfig{../figures/appendix/idI_elimination_new_component_hypergraph}
%             }
%             \]
%         \end{subfigure}
%     \caption{New consistent component introduction (top) and elimination (bottom) rule}
%     \label{fig:consistent_introduction}
% \end{figure}
\fi

@@

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.7\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix}
}
\]
\else
\begin{figure}[h!]
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix}
}
\]
\caption{Complete DPO square for the deletion rule}
\label{fig:cartesian_deletion_dpo}
\end{figure}
\fi
The target e-hypergraph is not exactly what we wanted, however, but we can use the auxiliary rule to remove this unlabelled edge.

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.7\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix_2}
}
\]
\else
\begin{figure}
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix_2}
}
\]
\caption{Complete DPO square for the deletion rule (after one more rule applied)}
\label{fig:cartesian_dpo_fix}
\end{figure}
\fi
% and then utilise the idempotence rule for the identity of the tensor unit
